### 2017-07-31 memocode_cr/conclusions.tex
MEMOCODE'17 camera ready

In this paper, we have proposed an approach to validation of VDM models by discharging proof obligations (POs) with SMT solvers. In our approach, Overture tool is used as the proof obligation generator, and the Python API of Z3 SMT solver is used for encoding and solving formulas of generated POs. Encoding strategies including steps of encoding as well as encoding patterns are demonstrated, and two case studies are reported. From our experiences, it is recognized that discharging POs using SMT solvers is efficient, and the counterexample generation of SMT solvers when the discharging fails can provide useful guidance for correcting inconsistency of VDM models. We also discussed other issues such as the applicability of encoding strategies considering the automation of our approach. As the future work, we are gathering and organizing more encoding patterns for challenging the automation issue.

>>>>

In this paper, we have proposed an approach to validation of VDM models by discharging proof obligations (POs) with SMT solvers. In our approach, Overture tool is used as the proof obligation generator, and the Python API of Z3 SMT solver is used for encoding and solving formulas of generated POs. Encoding strategies including steps of encoding as well as encoding patterns are demonstrated, and two case studies are reported. As initial results, it is recognized that discharging POs using SMT solvers is efficient, and the counterexample generation of SMT solvers is useful for correcting inconsistency of VDM models. Currently the encoding is done by hand. We are working on more concrete encoding patterns and automatic encoding tool for more case studies.
% We also discussed other issues such as the applicability of encoding strategies considering the automation of our approach. As the future work, we are gathering and organizing more encoding patterns for challenging the automation issue.

### 2017-07-31 memocode_cr/introduction.tex
MEMOCODE'17 camera ready

% General information about VDM
The Vienna Development Method (VDM)~\cite{Jones:1990:SSD:94062,Fitzgerald:2005:VDO:1044891,Fitz:PGLarson:2009book} is a formal method which supports modeling and analysis of software systems at various levels of abstraction. A VDM specification can be explicit, implicit, or both. An explicit specification is operational; it describes \emph{how} system components work. An implicit specification, on the other hand, is declarative; it describes \emph{what} system components should work. Software specifications in VDM models define functionalities explicitly or implicitly according to software requirements. VDM has a strong record of applications for design and specification of software systems in the industry~\cite{Larsen:2007:RIA:2227886.2227894,DBLP:conf/fm/KuritaCN08,DBLP:journals/ijsi/KuritaN09}. The well-known benefits of using VDM to describe software specifications are from the accuracy and unambiguity of VDM, which is common for other formal methods like B.

% Validation methods of VDM
To validate a VDM model, existing tools such as Overture~\cite{Larsen:2010:OII:1668862.1668864} and VDMTools provide static consistency check regarding the syntax and type constraints of VDM specifications. The semantic validation of VDM specifications is established by discharging proof obligations (POs)~\cite{AL:97:POGV} generated by the above tools. POs are logic formulas characterizing consistency of VDM specifications. Theorem provers have been used to discharge the generated POs. Testing (specification animation) by running a VDM model with an interpreter~\cite{Prehn:1991:LNCS551} is the alternative way of validation. This requires that the VDM model to be specified explicitly so that an interpreter can produce specific values for functions and operations.

% More about proof obligations and the motivation: solve as many as possible POs
Discharging POs of VDM models is essential to guarantee the consistency of the formal specifications. For a VDM model, the number of POs is usually large compared to the size of the model. For example, the sample VDM-SL model shown in Fig.~\ref{fig:module_sample} has less than thirty lines, and Overture generates nine POs for the model. Consequently, it is not uncommon to have hundreds of POs for large VDM models. All of them have to be discharged to guarantee consistency of software specifications. Manually discharging so many POs with theorem provers, however, can be a daunting task in practice. Furthermore, both erroneous VDM specifications and bad proof strategies induce un-dischargeable POs. Theorem provers provide little information when they fail to discharge POs. It is unclear whether a VDM specification is flawed or a better proof strategy is possible when POs cannot be discharged by theorem provers.

@@

In this paper, we apply SMT solvers to discharge POs of VDM specifications. Here we focus on VDM-SL models and the objective is to discharge POs of a VDM-SL model as many as possible through SMT solvers. In our approach, we encode each POs with its context information such as type and state constraints as a first-order formula, and then prove the PO by sending the formula to SMT solvers. The advantages of our approach are (1) the encoding involves only a few segments of a VDM specification; (2) the proof process is automated with SMT solvers; (3) If a PO's proof fails, a counterexample model is returned for further examination.

More specifically, in this approach, we chose the Overture tool as the PO generator. For the POs generated by the Overture tool, we encode and prove POs of VDM-SL models by solving SMT formulas with Python API of Z3~\cite{MB:08:ZSS}. Z3 is a popular SMT solver widely used in software verification and its Python API provides flexibility in constructing SMT formulas~\cite{BarFT-SMTLIB}. To encode POs, built-in VDM types such as sequences and maps must be characterized by theories in Z3. We conducted case studies and the results showed that our approach can efficiently discharge the significant part of POs of a VDM-SL model.
%  \footnote{POs generated by Overture and VDMTools are slightly different in the numbers and styles of POs}.

The structure of this paper is as follows: Section~\ref{section:vdm} briefly introduce VDM with a sample VDM-SL example; Section~\ref{section:proof-obligations} describes the POs of VDM and explains the POs generated by the Overture tool; Section~\ref{section:encode-strategy} describes our strategy of encoding VDM expressions with Python API of Z3; Section~\ref{section:case-studies} demonstrates case studies with discussions; Section~\ref{section:related-work} compares our approach to related work; Section~\ref{section:conclusions} concludes this paper and points out the future work.

>>>>

% General information about VDM
The Vienna Development Method (VDM)~\cite{Jones:1990:SSD:94062} is a formal method which supports modeling and analysis of software systems at various levels of abstraction. A VDM specification can be explicit, implicit, or both. An explicit specification is operational; it describes \emph{how} system components work. An implicit specification, on the other hand, is declarative; it describes \emph{what} system components should work. Software specifications in VDM models define functionalities explicitly or implicitly according to software requirements. 
% VDM has a strong record of applications for design and specification of software systems in the industry~\cite{Larsen:2007:RIA:2227886.2227894,DBLP:conf/fm/KuritaCN08,DBLP:journals/ijsi/KuritaN09}. The well-known benefits of using VDM to describe software specifications are from the accuracy and unambiguity of VDM, which is common for other formal methods like B.

% Validation methods of VDM
% To validate a VDM model, existing tools such as Overture~\cite{Larsen:2010:OII:1668862.1668864} and VDMTools~\cite{citeVDMtools} provide static consistency check regarding the syntax and type constraints of VDM specifications. The semantic validation of VDM specifications is established by discharging proof obligations (POs)~\cite{AL:97:POGV} generated by the above tools. POs are logic formulas characterizing consistency of VDM specifications. Theorem provers have been used to discharge the generated POs. Testing (specification animation) by running a VDM model with an interpreter~\cite{Prehn:1991:LNCS551} is the alternative way of validation. This requires that the VDM model to be specified explicitly so that an interpreter can produce specific values for functions and operations.

% More about proof obligations and the motivation: solve as many as possible POs
Discharging proof obligations (POs)~\cite{AL:97:POGV} of VDM models is essential to guarantee the consistency of the formal specifications. For a VDM model, the number of POs is usually large compared to the size of the model. 
% For example, the sample VDM-SL model shown in Fig.~\ref{fig:module_sample} has less than thirty lines, and Overture generates nine POs for the model. 
Consequently, it is not uncommon to have hundreds of POs for large VDM models. All of them have to be discharged to guarantee consistency of software specifications. Manually discharging so many POs with theorem provers, however, can be a daunting task in practice. Furthermore, both erroneous VDM specifications and bad proof strategies induce un-dischargeable POs. Theorem provers provide little information when they fail to discharge POs. It is unclear whether a VDM specification is flawed or a better proof strategy is possible when POs cannot be discharged by theorem provers.

@@

In this paper, we apply SMT solvers to discharge POs of VDM specifications. We focus on VDM-SL models and the objective is to discharge POs of a VDM-SL model as many as possible through SMT solvers. In our approach, we encode each POs with its context information such as type and state constraints as a first-order formula, and then prove the PO by sending the formula to SMT solvers. The advantages of our approach are (1) the encoding involves only a few segments of a VDM specification; (2) the proof process is automated with SMT solvers; (3) If a PO's proof fails, a counterexample model is returned for further examination.

We chose the Overture tool~\cite{Larsen:2010:OII:1668862.1668864} as the PO generator. For the POs generated by the Overture tool, we encode and prove POs of VDM-SL models by solving SMT formulas with Python API of Z3~\cite{MB:08:ZSS}. Z3 is a popular SMT solver widely used in software verification and its Python API provides flexibility in constructing SMT formulas~\cite{BarFT-SMTLIB}. To encode POs, built-in VDM types such as sequences and maps must be characterized by theories in Z3. We conducted case studies with manual encoding, and the results showed that our approach can efficiently discharge the significant part of POs of a VDM-SL model.
%  \footnote{POs generated by Overture and VDMTools are slightly different in the numbers and styles of POs}.

% The structure of this paper is as follows: Section~\ref{section:vdm} briefly introduce VDM with a sample VDM-SL example; Section~\ref{section:proof-obligations} describes the POs of VDM and explains the POs generated by the Overture tool; Section~\ref{section:encode-strategy} describes our strategy of encoding VDM expressions with Python API of Z3; Section~\ref{section:case-studies} demonstrates case studies with discussions; Section~\ref{section:related-work} compares our approach to related work; Section~\ref{section:conclusions} concludes this paper and points out the future work.

### 2017-07-31 memocode_cr/strategy.tex
MEMOCODE'17 camera ready

% Z3 introduction, API introduction
As mentioned in Section~\ref{section:introduction}, our approach encodes and proves VDM POs with the Z3 SMT solver. Z3 is one of the popular SMT solver wildly used in verification of programs and software systems. Z3 has APIs for major programming languages such as C, Java, and Python. The APIs provide a better way of constructing SMT formulas than SMT-LIB~\cite{BarFT-SMTLIB} because one can adopt the flow control characteristics like loop and if-then-else of programming languages to build and solve formulas in a flexible and smart way. In our approach, we chose Z3's Python API (Z3py) as the encoding environment.

@@

\begin{enumerate}
\item
Determine the context information of the PO: The context information is the definitions specified in the VDM model that are involved in the obligation. The context information may include types, functions, and operations. Note that if a type is involved, its type invariant should be included, too.
\item
Encode context information: this step encodes the context information determined by the first step and can be divided into several minor steps.
\begin{enumerate}
\item
For a VDM type in the context information, both the type itself and its invariant have to be encoded.
\item
For pre- and post-conditions of a function/operation specified in the VDM model, the encoding is needed if {\tt pre\_<name>} or {\tt post\_<name>} are used in the PO. For example, PO7 of the {\tt CMDS} module described in Section~\ref{section:proof-obligations} mentioned both precondition and postcondition, so both conditions shall be encoded.
\item
If a PO only considers a part of the specification of a type, a function, or an operations, it is only needed to encode the expressions that are directly used in the PO. In other words, if a type, a function, or an operation is not directly mentioned, for example, a variable mentioned with {\tt name:type} style, or a function mentioned with {\tt fun\_name(arguments)} style, there is no need to encode the full specification of the type, the function, or the operation.
\end{enumerate}
\item
Obligation negation and quantifier elimination
\begin{enumerate}
\item
Obligation negation: In model checking with SMT solvers, the assertion to be checked is usually negated, and the result of unsatmeans the proof of the assertion succeeds. For the result of sat which means the assertion can be violated, SMT solvers return a model as the counterexample that shows an example of how the assertion is violated. In this work, we also take negation on the PO to be discharged. However, the elimination of forall quantifier should be considered together when applying the negation to the PO.
\item
Elimination of forall quantifiers: POs of VDM usually have forall quantifiers. To solve the formulas with forall quantifiers in Z3, though Z3 accepts formulas with forall quantifiers, there are limitations applying forall quantifiers in Z3 so that by default we have to eliminate forall quantifiers before encoding. In this work, forall quantifier usually causes difficulties in encoding and solving when complex types are involved. Obligation negation described above may achieve the elimination of troublesome forall quantifiers. On the other hand, obligation negation should be applied carefully not to introduce troublesome forall quantifiers from exists quantifiers.
\end{enumerate}
\item
Encode the predicate of obligation and check for satisfiability.
\end{enumerate}

% More details of encoding types, operators, and expressions
VDM has rich types from natural number and quote to sequence and maps. There are no direct matching for all VDM types to the types in Z3. For example, to encode a natural number, we need to declare an integer and then add a constraint of greater and equal to zero because there is no natural number type in Z3. The enumeration type in Z3 can be used to encode simple quote types. For complex types like sequences, maps, or records, we may use uninterpreted functions or arrays as the base type and capture the characteristic and operators of the VDM types by adding corresponding constraints. At this point, there is not yet a complete and systematic way of encoding VDM types in Z3. What we can do is to choose the encoding carefully based on the types and expressions involved in a PO. Based on our case studies, there are patterns of encoding that are frequently used for corresponding expression styles in VDM-SL models.

% Demonstrate Z3Py encoding
For more details, the Z3py encoding firstly needs to include the Z3 module and declare a solver object. We can add constraints of a PO to the solver object and solve the formulas for proof.

\begin{mdframed}[roundcorner=5pt,shadow=true]
\begin{Verbatim}[fontsize=\small]
  from z3 import *
  s = Solver()
  i = Int('i')
  s.add(i>=0)
\end{Verbatim}
\end{mdframed}

The above Python code imports the z3 module and create a solver object, then declare an integer {\tt i} with a constraint that {\tt i} is greater and equal to zero. In this case, {\tt i} is the encoding an instance of type {\tt nat} in VDM. Note that we cannot add constraints on types but instances of types in Z3. This means that if there are several variables of the same VDM type that Z3 does not have, instances of the same number of variables have to be declared in Z3 with constraints for every instance added to the solver object. The steps and strategies of encoding discussed above will be demonstrated with details in Section~\ref{section:case-studies} through the report of case studies.

% Z3py encoding: taking negation and forall quantifier elimination
% Besides the encoding of VDM types, 
%  For example, if a VDM type is encoded as an uninterpreted function, the type cannot be used within forall quantifier. A simple way of quantifier elimination is negation the formula so the forall quantifier turns into exists quantifier which can be omitted in SMT formulas usually.

>>>>

% Z3 introduction, API introduction
In this work, we encode and prove VDM POs with the Z3 SMT solver. Z3 is one of the popular SMT solver wildly used in verification of programs and software systems. Z3 has APIs for major programming languages such as C, Java, and Python. The APIs provide a better way of constructing SMT formulas than SMT-LIB~\cite{BarFT-SMTLIB} because one can adopt the flow control characteristics like loop and if-then-else of programming languages to build and solve formulas in a flexible and smart way. In our approach, we chose Z3's Python API (Z3py) as the encoding environment.

@@

\begin{enumerate}
\item
Determine the context information of the PO.
% : The context information is the definitions specified in the VDM model that are involved in the obligation. The context information may include types, functions, and operations. Note that if a type is involved, its type invariant should be included, too.
\item
Encode context information.
% : this step encodes the context information determined by the first step and can be divided into several minor steps.
\begin{enumerate}
\item
For a VDM type in the context information, both the type itself and its invariant have to be encoded.
\item
If pre/post-conditions of functions/operations are used in a PO in the form of {\tt pre\_<name>} or {\tt post\_<name>}, encoding of pre/post-conditions is needed.
%  are used in the PO. For example, PO7 of the {\tt CMDS} module described in Section~\ref{section:proof-obligations} mentioned both precondition and postcondition, so both conditions shall be encoded.
\item
If a PO only considers a part of the specification of a type, a function, or an operation, it is only needed to encode the expressions that are directly used in the PO.
%  In other words, if a type, a function, or an operation is not directly mentioned, for example, a variable mentioned with {\tt name:type} style, or a function mentioned with {\tt fun\_name(arguments)} style, there is no need to encode the full specification of the type, the function, or the operation.
\end{enumerate}
\item
Obligation negation and quantifier elimination.
\begin{enumerate}
\item
Obligation negation: In model checking with SMT solvers, the assertion to be checked is usually negated, and the result of unsat means the proof of the assertion succeeds. For the result of sat which means the assertion can be violated, SMT solvers return a model as the counterexample that shows an example of how the assertion is violated. In this work, we also negate the PO to be discharged. However, the elimination of universal quantifier should be considered together when applying the negation to the PO.
\item
Elimination of universal quantifiers: POs of VDM are usually with universal quantifiers. Though Z3 accepts formulas with universal quantifiers, there are limitations applying universal quantifier in Z3. Therefore, we eliminate universal quantifiers in the predicate of a PO before encoding. Obligation negation described above also achieves the elimination of universal quantifiers. 
% On the other hand, for POs with existential quantifiers, obligation negation is not needed for not introducing universal quantifiers. 
However, for POs with both universal and existential quantifiers, or multiple universal quantifiers, it should be avoid introducing troublesome universal quantifiers, i.e., universal quantifiers on sequences, sets, and maps, from exists quantifiers.
\end{enumerate}
\item
Encode the predicate of obligation and check for satisfiability.
\end{enumerate}

% More details of encoding types, operators, and expressions
VDM has rich types from natural number and quotes to sequences and maps. There are no direct matching for all VDM types to the types in Z3. For example, to encode a natural number, we need to declare an integer and then add a constraint of greater and equal to zero because there is no natural number type in Z3. The enumeration type in Z3 can be used to encode simple quote types. For complex types like sequences, maps, or records, we may use uninterpreted functions or arrays as the base type and capture the characteristic and operators of the VDM types by adding corresponding constraints. At this point, there is not yet a complete and systematic way of encoding VDM types in Z3. What we can do is to choose the encoding carefully based on the types and expressions involved in a PO. Based on our case studies, there are patterns of encoding that are frequently used for corresponding expression styles in VDM-SL models.
